---
title: Comparison of models for credit risk purposes - logistic regression vs random
  forest
author: "Katarzyna Kryńska"
date: "5 07 2020"
output: 
  rmdformats::material:
    highlight: tango
    cards: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rmdformats)
library(corrplot)
library(RColorBrewer)
library(kableExtra)
library(tidyverse)
```


```{r do_not_run, eval=F, include=F}
# load data from OpenML and save to to a local directory
library(OpenML)
credit_OpenML <- getOMLDataSet(data.id = 31)
df <- credit_OpenML$data
save(df, file="df.rda")
```


# Abstract


# Dataset description

```{r load_df, eval=F, include=F}
load("load/df.rda")

# all variables are either factor or numeric, so that's good
# BUT class has levels 1-2 and not 0-1, which might be a problem?
# safer to change it into 0-1 values, where 0 is good and 1 is bad
df$class <- as.numeric(df$class)-1
```

```{r summary_df, eval=F, include=F}
# PRE-ANALYSIS OF DATA - SUMMARY, DESCRIPTION

## summary
#summary(df)

## Desc
#library(DescTools)
#Desc(df)

## are there any NAs? ----> NO
colSums(is.na(df))

## str
str(df)

# df$class <- factor(as.numeric(df$class) - 1, levels = c(0,1), labels = c("good", "bad"))

```

```{r merge_factor_variables, include=F, eval=F}

## Merge some factor variables

# credit_history
# merge "all paid" with "no credits/all paid"
new_labels <- levels(df$credit_history)[!levels(df$credit_history) %in% c("all paid")]
new_levels <- c(1, 3, 4, 5)
df$credit_history <- factor(ifelse(as.numeric(df$credit_history) %in% c(1,2), 1, df$credit_history), labels=new_labels)

# purpose
# 6 "repairs" dorzucam do 11 "other"
# 5 "domestic appliance" do 3 "furniture/equipment"
# 9 "retraining" do 7 "education"

new_levels <- c(1:11)[!c(1:11) %in% c(6, 5, 9)]
new_labels <- levels(df$purpose)[new_levels]
new_labels[5] <- "education/retraining"

df$purpose <- as.numeric(df$purpose)
df$purpose <- ifelse(df$purpose %in% c(6,11), 11, df$purpose)
df$purpose <- ifelse(df$purpose %in% c(5,3), 3, df$purpose)
df$purpose <- ifelse(df$purpose %in% c(9,7), 7, df$purpose)
df$purpose <- factor(df$purpose, levels=new_levels, labels=new_labels)

# existing_credits
df$existing_credits <- ifelse(df$existing_credits %in% c(3,4), 3, df$existing_credits)
new_levels <- c(1:3)
new_labels <- c("1", "2", ">2")
df$existing_credits <- factor(df$existing_credits, levels=new_levels, labels=new_labels)

```

```{r DISCRETIZATION, eval=F, include=F}
# DISCRETIZATION - FINE CLASSING

## division variables into FACTORS and NUMERIC variables
base <- df[, colnames(df) != "class"]
base.n<-base[,sapply(base, is.numeric)]
base.f<-base[,sapply(base, is.factor)]

## percentiles
percentile<-apply(X=base.n, MARGIN=2, FUN=function(x) round(quantile(x, seq(0.1,1,0.1), na.rm=TRUE),2))

## unique values per column
unique<-apply(base.n, MARGIN=2, function(x) length(unique(x)))

## selecting only columns with more than 10 unique levels
## binarization for numeric variables with >10 unique values
numeric<-colnames(base.n[which(unique>=10)])
for (m in numeric){
  base.n[,paste(m,"_fine", sep="")]<-cut(
    x=as.matrix(base.n[m]), 
    breaks=c(-Inf,unique(percentile[,m])), 
    labels =c(paste("<=",unique(percentile[,m])))
    ) 
}

## columns with <10 levels will be transformed to factors
num_as_fact<-colnames(base.n[which(unique<10 & unique>1 )])
base.f[,num_as_fact]<-lapply(base.n[,num_as_fact],as.factor)
base.n <- base.n[,!names(base.n) %in% num_as_fact]
```

```{r WOE_calc, eval=F, include=F}
# Calculation of Weight of Evidence (WoE) using smbinning

## in smbinning 0 means bad, --> reverting definition of gb flag
base.n$class_woe<-(1-df$class)
base.n$class <- df$class

base.f$class_woe <- base.n$class_woe
base.f$class <- base.n$class

## Choice of vars to be analysed
names.n<-colnames(base.n[,!names(base.n) %in% c(
  colnames(base.n[ ,grepl(pattern="_fine" , x=names(base.n))]),
  "class",
  "class_woe")])
names.f<-colnames(base.f[,!names(base.f) %in% c("class","class_woe")])

## List for WOE - preallocation
WOE <- list()

## Dataframe for Information Value
IV <- data.frame(VAR=character(), IV=integer())
```

```{r WoE_numeric.pdf, eval=F, include=F}
# ANALYSIS OF NUMERIC VARIABLES
library(smbinning)

# Storing charts in a pdf file
pdf(file="WoE_numeric.pdf",paper="a4")


#Set up progress bar
pb<- txtProgressBar(min = 0, max = length(names.n), style = 3)

for (i in names.n){
  
  # rows and columns at a chart
  par(mfrow=c(2,2))
  
  # smbinning.custom(df, y, x, cuts)
  # df - data frame
  # y - GB flag
  # x - risk factors
  # cuts - cut-offs
  results<- smbinning.custom(df=base.n, y="class_woe", x=i, cuts=unique(percentile[,i]))
  
  # BOXPLOT
  boxplot(base.n[,i]~base.n$class, horizontal=T, frame=F, col="lightgray",main="Distribution") 
  mtext(i,3) 
  # Frequency plot 
  smbinning.plot(results,option="dist",sub=i)
  # Bad rate fractions
  smbinning.plot(results,option="badrate",sub=i) 
  # WoE
  smbinning.plot(results,option="WoE",sub=i)

  # IV row binding
  IV<-rbind(IV, as.data.frame(cbind("VAR"=i, "IV"=results$ivtable[results$ivtable$Cutpoint=="Total", "IV"])))
  
  # Saving data 
  d<-results$ivtable[,c("Cutpoint","WoE","PctRec")]
  # Total row removal  
  d<-d[d$Cutpoint!="Total",]
  # Ordering wrt WoE - for gini etc. calculation
  d<-d[with(d, order(d$WoE)),]
  # Id 
  d$numer<-11:(nrow(d)+10)
  # Saving WoE in list 
  WOE[[i]]<-d
  
  #Update progess bar  
  setTxtProgressBar(pb,  min(grep(i, names.n)))
} 

close(pb)
dev.off()

# Reseting charts options
resetPar <- function() {
  dev.new()
  op <- par(no.readonly = TRUE)
  dev.off()
  op
}

resetPar()

# Insights:
# 1) duration - almost linear relationship between credit duration and probability of default
# 2) credit_amount - looks like a non-linear relationship, mid credit_amounts have the smallest pod,
# while the smallest and especially biggest credits are more prone to default
# 3) age - relationship not clear, the risk seems to be dropping with bigger age, but people (52;75] seem
# riskier than people (36;52]
```

```{r WoE_factor.pdf, include=F, eval=F}
str(base.f)

# PDF
pdf(file="WoE_factor.pdf",paper="a4")

# progress bar
pb<- txtProgressBar(min = 0, max = length(names.f), style = 3)

for (i in names.f){

  #neede to join results for factors and numerics
  base.f[,paste(i,"_fine", sep="")]<-base.f[,i]
    
  par(mfrow=c(2,2))
  
  results <- smbinning.factor(df=base.f, y="class_woe", x=i, maxcat=length(unique(base.f[,i])))
  # df - data frame
  # y - GB
  # x - variable 
  # maxcat - maximal accepted number of categories

  smbinning.plot(results,option="dist",sub=i) 
  smbinning.plot(results,option="badrate",sub=i) 
  smbinning.plot(results,option="WoE",sub=i)
  
  IV<-rbind(IV, as.data.frame(cbind("VAR"=i, "IV"=results$ivtable[results$ivtable$Cutpoint=="Total", "IV"])))

  d<-results$ivtable[,c("Cutpoint","WoE","PctRec")]
 
  d<-d[d$Cutpoint!="Total",]
 
  d<-d[with(d, order(d$WoE)),]
  
  d$numer<-11:(nrow(d)+10)
  
  WOE[[i]]<-d
  
  setTxtProgressBar(pb,  min(grep(i, names.f)))
} 
close(pb)
dev.off()

```

```{r variables_quality_gini_iv, eval=F, include=F}
base <- cbind(base.f[,!names(base.f) %in% c("class_woe", "class")], base.n)
library(forcats)
library(pROC)

# progress bar
pb<- txtProgressBar(min = 0, max = length(names(WOE)), style = 3)
stats<-cbind(IV, Gini=NA, miss=NA)
l<-"checking_status"

for (l in names(WOE)){

  # objects need for Gini calculation
  variable<-base[,c("class_woe", paste(l,"_fine", sep=""))]
  woe<- WOE[[l]][c("Cutpoint", "WoE")]
  
  # Cutpoint values change if Factor
  if (is.character(woe$Cutpoint)==TRUE) { 
    woe$Cutpoint<-as.factor(gsub("= '|'", "", woe$Cutpoint))
    woe$Cutpoint<-as.factor(woe$Cutpoint)
  }
  
  # Var levels and WoE merging
    # merge()
    # x,y - tables
    # by.x .y key
    # all.x=T - left join
  
  dat_temp<-merge(variable, woe, by.x=paste(l,"_fine", sep=""), by.y="Cutpoint", all.x=T)

 # name change
  colnames(dat_temp)[which(names(dat_temp) == "WoE")] <- paste(l, "_woe", sep="")  
  
  # adding WoE var to original dataset
  base<-merge(base, woe, by.x=paste(l,"_fine", sep=""), by.y="Cutpoint", all.x=T)
  colnames(base)[which(names(base) == "WoE")] <- paste(l, "_woe", sep="")
  
  # gini claculation
  #   gini = 2*AUROC-1
  #   auc() function for AUROC calculation
  #   ?pROC
  
  gini<- c(2*auc(dat_temp$class_woe,dat_temp[,paste(l, "_woe", sep="") ])-1)
  
  stats[stats$VAR==l, "Gini"]<-gini
  
  miss<-1-c(nrow(dat_temp[dat_temp[,paste(l,"_fine", sep="")]!='Missing', ])/nrow(dat_temp))
  
  stats[stats$VAR==l, "miss"]<-miss
  
  setTxtProgressBar(pb,  min(grep(l, names(WOE))))
  
}

close(pb)

write.csv(stats, "stats.csv")

```

```{r variables_quality_kendall_correlation, include=F, eval=F}

# only _woe variables
var_to_check<-colnames(base)[grep("_woe", colnames(base))]
# exclusion of class_woe
var_to_check<-var_to_check[-1]
base_kor<-base[,var_to_check]

# cor()
# x- base
# method - type of corr coeff
# use - what to do if missings?; "pairwise" - deleting observations only if missing in considered vars,
kendall <-cor(base_kor, method = "kendall",use="pairwise")
save(kendall, file="load/kendall.Rdata")
```

```{r variables_selection, include=F, eval=F}

corr_list<-data.frame()

# ordering correlation matrix wrt Gini (rows & columns)
#woe var creation
stats$VARW<-paste(stats$VAR,"_woe",sep="")
# merging statistics & correlation results
stat<-merge(x = stats, y = kendall, by.x = "VARW",by.y="row.names",all.y=T)
#ordering wrt Gini - descending
stat<-stat[ order(-stat[,"Gini"]), ]
#changing row names as variables names
row.names(stat)<-stat[,1]
#only correlation matrix
stat<-stat[6:length(stat[1,])]
#vector of names
cols<-row.names(stat[,])
#coloum ordering wrt Gini
stat<-stat[cols]

#if no correlation (missings) then small value of correlation assigning, 
#that won't delete var from the analysis
temp_k<-stat
temp_k<-replace(temp_k, is.na(temp_k), 0.000001000)
# 
# loop reapet()
# it works till stop

# install.packages("plyr")
library(plyr)
threshold<-0.5

repeat{
  #even though this condistion is at the beggining, it has rather final character 
  # condition to stop a loop is reduction to NULL object
  # it may happend when in the last loop all variables satisfy a condition
  # > threshold 
  if (length(temp_k)<1) {
    break
  }
  
  # selection of all variables that correlation with a first variable is higher than threshold
  # NOTE: correlation coefficient between variable and  itself is equal 1 so it wuold be taken into data frame
  row_k<-abs(temp_k[1,]) > threshold
  row_k[1]<-TRUE
  
  # condition checkig wheter we are at the end of the analysis
  # if not then vars that > threshold is satisfied are taken
  # in yes the final variable is taken
  if(length(row_k)>1){
    row_k2<-row_k[,row_k]
  }else{row_k2<-row_k}
  
  # from temp_k all rows kept in row_k2 are deleted
  
  temp_k<-as.data.frame(temp_k[!t(row_k),!row_k])
  
  #if only one variable remains then R would reduce DF to vector and names for rows and columns have to be defined
  if(length(temp_k)==1){ 
    colnames(temp_k)<- dimnames(row_k)[[2]][-which(row_k)]
    rownames(temp_k)<- dimnames(row_k)[[2]][-which(row_k)]
  }
  
  
  #creation of vecotr with variables that satisfy > threshold
  if(length(row_k2)==1) {
    variable<-row.names(row_k)
    namess<-as.data.frame(variable)
  }
  
  if(length(row_k2)>1) {
    row_k2<-row_k2[2:length(row_k2)]
    row_k2<-as.data.frame(t(row_k2))
    namess<-colnames(row_k2)
    namess<-as.data.frame(t(namess))
    row.names(namess) <- row.names(row_k)
    variable<-row.names(row_k)
    variable<-as.data.frame(variable)
    namess<-merge(variable,namess)
  }
  
  
  # adding to corr_list row with all correlated vars with the variable
  # rbind.fill() - filling missing columns with NA
  
  corr_list<-rbind.fill(corr_list,namess)
  
  # condition to leave a loop - when last variable left
  
  if(length(temp_k)==1){ 
    variable<-dimnames(row_k)[[2]][-which(row_k)]
    namess<-as.data.frame(variable)
    corr_list<-rbind.fill(corr_list,namess)
    break}
  
}

# merging of data with stats and correlations
all<-merge(x = stats, y = corr_list, by.x = "VARW",by.y="variable",all.y=T)
# ordering by Gini
all<-all[ order(-all[,"Gini"]), ]
# exclusion of variables with gini below 0.1
all<-all[all$Gini>0.1,]

base_coarse<-base[-grep("_fine",colnames(base))]
kols<-c(as.character(all$VAR),"class","class_woe")
base_coarse<-base_coarse[,kols]

```

```{r sample_split, include=F, eval=F}

#drawing with stratification  
library(caTools)

# two subsamples, one for each subsample
# list with number of strata dimensions
# f - strata
d_split<-split(x=base_coarse, f=base_coarse$class)

# table train i test
train = list()
test=list()


# for loop for each strata

for (i in 1:length(d_split)){
  
  set.seed(1916)

  # splitting vector for i-th strata
  assign(
    paste("sample",i,sep="_"),sample.split(d_split[[i]][,1], SplitRatio = .7))
  
  # training set for i-th starta
  assign(
    paste("train",i,sep="_"),subset(d_split[[i]],  get(paste("sample",i,sep="_")) == TRUE))
  
  # validation set for i-th starta
  assign(
    paste("test",i,sep="_"),subset(d_split[[i]], get(paste("sample",i,sep="_")) == FALSE))
  
  train[[i]]<-get(paste("train",i,sep="_"))
  test[[i]]<-get(paste("test",i,sep="_"))
}

# row binding of lists train & test
train = do.call(rbind, train)
test = do.call(rbind, test)

# removing unnecessery objects
rm( test_1, test_2
    , train_1, train_2
    , sample_1, sample_2)

mean(train$class)
mean(test$class)

save(train,file="train.RData")
save(test,file="test.RData")

```

```{r WoE_coarse_numeric.pdf, include=F, eval=F}
load("load/test.RData")
load("load/train.RData")

columns <- colnames(train)[-which(names(train) %in% c("class","class_woe"))]
columns.n<-columns[sapply(train[,columns], is.numeric)]

source("ivmult.r")
library(ggplot2)

# progress bar
waitBar <- txtProgressBar(min = 0, max = length(columns.n), style = 3)

temp<-data.frame(VAR=character(), VALUE=integer())
stats<-data.frame(VAR=character(), IV=integer(),Gini=integer(), MISS=integer(),IVw=integer(),Giniw=integer(), MISSw=integer())

#cut offs for testing data
cut_offs<-data.frame(VAR=character(), cuts=integer())

current_var<-columns.n[2]

pdf("WoE_coarse_numeric.pdf")

for (current_var in columns.n){
  
  par(xpd = T, mar = par()$mar, mfrow=c(2,2))
  
  # smbinning using decision tree
  # Optimal Binning categorizes a numeric characteristic into bins for ulterior usage in scoring modeling. 
  # This process, also known as supervised discretization, utilizes Recursive Partitioning to categorize the numeric characteristic.
  # The especific algorithm is Conditional Inference Trees which initially excludes missing values (NA) to compute the cutpoints, adding them back later in the process for the calculation of the Information Value.
  
  result<-smbinning(train[,c("class_woe", current_var)], 
                    y="class_woe", 
                    x=current_var, 
                    p=0.05)
  # for age
  if(length(result)<=1){
    # only for age
    age_cut <- c(25, 33)
    result<- smbinning.custom(df=train, y="class_woe", x=current_var, cuts=age_cut)
  }
  
  
  if(length(result)>1){
    points<-list(result$cuts)
    cut_offs<-rbind(cut_offs,as.data.frame(cbind("VAR"=current_var,"cuts"=points)))
    
    IV<-result$ivtable
    
    #coarsed variables  
    train[,paste(current_var, "_coarse", sep="")]<- cut(train[,current_var], 
                                                        breaks=c(-Inf,unique(result$cuts),Inf), 
                                                        labels=c(paste("<=", unique(result$cuts)),"<= Inf"),
                                                        include.lowest = T)
    train[,paste(current_var, "_coarse", sep="")]<-fct_explicit_na(train[,paste(current_var, "_coarse", sep="")], na_level="Missing")
    
    test[,paste(current_var, "_coarse", sep="")]<- cut(test[,current_var],
                                                       breaks=c(-Inf,unique(result$cuts),Inf), 
                                                       labels=c(paste("<=", unique(result$cuts)),"<= Inf"),
                                                       include.lowest = T)
    test[,paste(current_var, "_coarse", sep="")]<-fct_explicit_na(test[,paste(current_var, "_coarse", sep="")], na_level="Missing")
    
    
    IVw<-sum(iv.mult(test,"class",vars=paste(current_var,"_coarse",sep=""))[[1]][,"miv"])
    IV$Cutpoint<-ifelse(grepl(">",IV$Cutpoint)==T,"<= Inf",IV$Cutpoint)
    
    #addig WoE variable
    train<-merge(train,IV[,c("Cutpoint", "WoE")],by.x= paste(current_var, "_coarse", sep=""), by.y="Cutpoint", all.x=T, sort=F)
    colnames(train)[which(names(train) == "WoE")] <- paste(current_var, "_woe", sep="") 
    
    test<-merge(test,IV[,c("Cutpoint", "WoE")],by.x= paste(current_var, "_coarse", sep=""), by.y="Cutpoint", all.x=T, sort=F)
    colnames(test)[which(names(test) == "WoE")] <- paste(current_var, "_woe", sep="")
    
    #GINI
    gini<- 2*auc(train$class_woe,train[,paste(current_var, "_woe", sep="") ])-1
    giniw<- 2*auc(test$class_woe,test[,paste(current_var, "_woe", sep="") ])-1
    # % missings
    miss<-1-nrow(train[!(is.na(train[,current_var])),])/nrow(train)
    missw<-1-nrow(test[!(is.na(test[,current_var])),])/nrow(test)
    #stats<-rbind(stats, as.data.frame(cbind("VAR"=current_var, "IV"=IV[IV$Cutpoint=="Total", "IV"],"Gini"=gini,  "MISS"=miss,"IVw"=IVw, "Giniw"=giniw,  "MISSw"=missw)))
    
    plot<-IV[!is.na(IV$WoE) & IV$Cutpoint!="Total",]
    plot$WoE<-ifelse(plot$WoE==Inf, 8, ifelse(plot$WoE==-Inf, -8,plot$WoE))
    g<-barplot(plot$WoE,names.arg=plot$Cutpoint, cex.names=0.5, main=current_var, xaxt="n")
    print(g)
    axis(1,g,plot$Cutpoint, tick=F, las=2, cex.axis=0.7)
    text(g,plot$WoE, labels=paste("WoE=",format(plot$WoE,digits=1)," \n Fill=",round(plot$PctRec,3)*100, "%", sep=""), xpd = T, col = "black", pos=3, cex=0.7)
    
    # stability
    freq.test<-as.data.frame(table(test[,paste(current_var, "_coarse", sep="")])/ nrow(test))
    colnames(freq.test)<-c("Cutpoint", "PctRec")
    freq.train<-plot[,c("Cutpoint","PctRec")]
    freq<-merge(freq.train,freq.test, all.x=T, by="Cutpoint")
    colnames(freq)<- c("Cutpoint","Train","Test")
    rownames(freq)<-freq[,1]
    freq<-freq[,-1]
    
    stacked<-barplot(as.matrix(freq), beside=F, main=current_var)
    print(stacked)
    legend("topright", legend=row.names(freq), cex=0.6, bty="o",pch=21)
    
    smbinning.plot(result,option="dist",sub=current_var)
    
    smbinning.plot(result,option="badrate",sub=current_var) 
    
  } 
  
  setTxtProgressBar(waitBar,  min(grep(current_var, columns.n)))
}
dev.off()

```

```{r woeBinning, include=F, eval=F}
library(woeBinning)
columns.f <- columns[sapply(train[,columns], is.factor)]
woe.bin_f <- woe.binning(train, 
                         target.var="class", 
                         pred.var=columns.f, 
                         min.perc.total=0.05,
                         min.perc.class=0.02, 
                         stop.limit=0.1)
# min.perc.total =	
#   For numeric variables this parameter defines the number of initial classes before any merging is applied. For example min.perc.total=0.05 (5%) will result in 20 initial classes. 
#   For factors the original levels with a percentage below this limit are collected in a ‘miscellaneous’ level before the merging based on the min.perc.class and on the WOE starts. 
#   Accepted range: 0.01-0.2; default: 0.05.

# min.perc.class =
#   If a column percentage of one of the target classes within a bin is below this limit (e.g. below 0.01=1%) then the respective bin will be joined with others. In case of numeric variables adjacent predictor classes are merged. 
#   For factors respective levels (including sparse NAs) are assigned to a ‘miscellaneous’ level. 
#   Accepted range: 0-0.2; default: 0, i.e. no merging with respect to sparse target classes is applied.

# stop.limit	
# Stops WOE based merging of the predictor's classes/levels in case the resulting information value (IV) decreases more than x% (e.g. 0.05 = 5%) compared to the preceding binning step. stop.limit=0 will skip any WOE based merging. 
# Increasing the stop.limit will simplify the binning solution and may avoid overfitting. 
# Accepted range: 0-0.5; default: 0.1.

woe.binning.plot(woe.bin_f, multiple.plots=F)
woe.binning.table(woe.bin_f)
train<-woe.binning.deploy(train, woe.bin_f,add.woe.or.dum.var='woe')
test<-woe.binning.deploy(test, woe.bin_f,add.woe.or.dum.var='woe')


# I should unify the names of binned/coarse variables and woe
library(stringr)
## woe.credit_history.binned --------> credit_history_woe
pattern <- "woe.\\w+.binned"
idx <- str_detect(colnames(train), regex(pattern))
new_str <- str_extract(colnames(train),pattern) %>% str_remove("woe.") %>% str_remove(".binned") %>% str_c("_woe")
colnames(train) <- ifelse(idx, new_str, colnames(train))
colnames(test) <- ifelse(idx, new_str, colnames(train))
## checking_status.binned ----------> "checking_status_coarse
pattern <- "\\w+.binned"
idx <- str_detect(colnames(train), regex(pattern))
new_str <- str_extract(colnames(train),pattern) %>% str_remove(".binned") %>% str_c("_coarse")
colnames(train) <- ifelse(idx, new_str, colnames(train))
colnames(test) <- ifelse(idx, new_str, colnames(train))

str(test)

```

```{r WOE_coarse_all.pdf, include=F, eval=F}

columns.coarse <- str_extract(colnames(train),"\\w+_coarse")[str_detect(colnames(train), "\\w+_coarse")]

# progress bar
waitBar<- txtProgressBar(min = 0, max = length(columns.coarse), style = 3)

resetPar <- function() {
  dev.new()
  op <- par(no.readonly = TRUE)
  dev.off()
  op
}

pdf("WOE_coarse_all.pdf")

current_variable <- "credit_amount"
current_variable<-columns.coarse[1]

for (current_variable in columns.coarse){
  
  par(xpd = T, mfrow=c(2,2))
  result<-smbinning.factor(train[,c("class_woe", current_variable)], y="class_woe", x=current_variable)

  points<-list(result$cuts)
  cut_offs<-rbind(cut_offs,as.data.frame(cbind("VAR"=current_variable,"cuts"=points)))
  
  IV<-result$ivtable
  IVw<-sum(iv.mult(test,"class",vars=current_variable)[[1]][,"miv"])
  IV$Cutpoint<-as.factor(gsub("= |'","",IV$Cutpoint))

  #GINI
  gini<- 2*auc(train$class_woe,train[,gsub("_coarse","_woe",current_variable) ])-1
  giniw<- 2*auc(test$class_woe,test[,gsub("_coarse","_woe",current_variable)])-1
  
  miss<-1-nrow(train[!(is.na(train[,current_variable])),])/nrow(train)
  missw<-1-nrow(test[!(is.na(test[,current_variable])),])/nrow(test)
  stats<-rbind(stats, as.data.frame(cbind("VAR"=current_variable, "IV"=IV[IV$Cutpoint=="Total", "IV"],"Gini"=gini,  "MISS"=miss,"IVw"=IVw, "Giniw"=giniw,  "MISSw"=missw)))
  
  #plot WoE
  plot<-IV[!is.na(IV$WoE) & IV$Cutpoint!="Total",]
  plot$WoE<-ifelse(plot$WoE==Inf, 8, ifelse(plot$WoE==-Inf, -8,plot$WoE))
  g<-barplot(plot$WoE,names.arg=plot$Cutpoint, cex.names=0.5, main=current_variable, xaxt="n")
  print(g)
  axis(1,g,plot$Cutpoint, tick=F, las=2, cex.axis=0.7)
  text(g,plot$WoE, labels=paste("WoE=",format(plot$WoE,digits=1)," \n Fill=",round(plot$PctRec,3)*100, "%", sep=""), xpd = T, col = "black", pos=3, cex=0.7)
  
  # stability
  
  freq.test<-as.data.frame(table(test[,current_variable])/ nrow(test))
  colnames(freq.test)<-c("Cutpoint", "PctRec")
  freq.train<-plot[,c("Cutpoint","PctRec")]
  freq<-merge(freq.train,freq.test, all.x=T, by="Cutpoint")
  colnames(freq)<- c("Cutpoint","Train","Test")
  rownames(freq)<-freq[,1]
  freq<-freq[,-1]
  
  stacked<-barplot(as.matrix(freq), beside=F, main=current_variable)
  print(stacked)
  
  smbinning.plot(result,option="dist",sub=current_variable)
  
  smbinning.plot(result,option="badrate",sub=current_variable) 

setTxtProgressBar(waitBar,  min(grep(current_variable, columns.f)))
}

dev.off()
warnings()
par(resetPar()) 

#Saving -----------------------------------

#write.csv(stats, file="stats_final.csv")
save(train,file="load/train_final.Rdata")
save(test,file="load/test_final.Rdata")
```

Dataset that was used in this research is a well-known *credit-g* dataset, available at OpenML [https://www.openml.org/d/31](https://www.openml.org/d/31). Original dataset contains 1 target value, *class*, which describes if the customer is good or bad, and 21 attributes:

1. *checking_status* - Status of existing checking account, in Deutsche Mark.
2. *duration* - Duration in months
3. *credit_history* - Credit history (credits taken, paid back duly, delays, critical accounts)
4. *purpose* - Purpose of the credit (car, television,...)
5. *credit_amount* - Credit amount
6. *savings_status* - Status of savings account/bonds, in Deutsche Mark.
7. *employment* - Present employment, in number of years.
8. *installment_commitment* - Installment rate in percentage of disposable income
9. *personal_status* - Personal status (married, single,...) and sex
10. *other_parties* - Other debtors / guarantors
11. *residence_since* - Present residence since X years
12. *property_magnitude* - Property (e.g. real estate)
13. *age* - Age in years
14. *other_payment_plans* - Other installment plans (banks, stores)
15. *housing* - Housing (rent, own,...)
16. *existing_credits* - Number of existing credits at this bank
17. *job* - Job
18. *num_dependents* - Number of people being liable to provide maintenance for
19. *own_telephone* - Telephone (yes,no)
20. *foreign_worker* - Foreign worker (yes,no)

The datasets contains data for 1000 customers, of which 700 are good and 300 are bad ones. This means that the data is not much imbalanced and there is enough number of bad clients to estimate/train a reliable model.

To build a scorecard model based on logistic regression, coarse classing of the data is needed. Firstly, continuous variables were binned into factor variables. The initial division into bins was made with respect to percentiles. In case of some variables which were originally factors, merging similar categories was done. 

In the end, we ended up with 20 factor variables. Then, we checked their correlation, using Kendall correlation measure. This is important in order not to introduce colinearity into our model. The result is plotted below:

```{r plot_kendall_corr, out.width = '80%', fig.align='center', fig.asp = .80}

load("load/kendall.Rdata")
corrplot(kendall, type="upper", 
         order="hclust",
         col=brewer.pal(n=8, name="BrBG"),
         tl.cex=0.55,
         tl.col="brown",
         tl.offset=1,
         tl.srt=45,)

```

It seems like there will not be any problems with colinearity in the model. We might only consider removing either *credit_history* or *existing_credits*, as their correlation is around 50%.

Then, we calculated some variables quality statistics, like Gini and Information Value.

```{r print_variables_quality_table}
quality_table <- read.csv("stats.csv")
quality_table <- quality_table %>% arrange(desc(Gini)) %>% subset(select=-c(X, miss))
quality_table$"No." <- c(1:20)
quality_table <- quality_table[c(4, 1, 2, 3)]
kable(quality_table, digits = 4) %>% kable_styling(bootstrap_options = c("striped", "hover","condensed", "responsive"))
```

We decided to include in our final dataset only these variables, for which Gini value is bigger than 10%. In the end we ended up with 11 explanatory variables.

To divide the dataset into *train* and *test* samples, we used stratified sampling. This way the ratio between number of good and bad clients is the same for *train* and *test* samples. 

Then, using *train* sample, we performed supervised discretization. We used Optimal Binning to categorise numeric variables into bins for scoring modeling. Only for age bins were not assigned by the algorithm, so they were created manually, based on WoE of previously created percentiles. In case of factor variables, they were automatically merged into groups with the use of woeBinning package. 




# Logistic regression


# Random forest


# Summary
